---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# ggmissing

ggplot does not display missing data and currently provides a warning message and omits the missing data.`ggmissing` aims to enable ggplot to display missingness. 

`ggmissing` is part of a larger plan for a set of tidy-verse packages focussing on how to tidy, transform, visualise, model, and communicate missing data.

It is still very much under development, but will see more active development over the next 6 months.

## What does it do?

Plotting missing data might sound a little strange - how do you visualise something that is not there? In the past, GGobi and Manet have provided methods of visualising missingness, with one approach being to replace "NA" values with values 10% lower than the minimum value in that variable.

To illustrate, let's create some toy data using the wakefield package.

```{r}
library(wakefield)
library(ggmissing)
# devtools::install_github("njtierney/ggmissing")
library(ggplot2)
library(dplyr)

df <- 
  r_data_frame(
  n = 30,
  id,
  race,
  age,
  sex,
  hour,
  iq,
  height,
  died,
  Scoring = rnorm,
  Smoker = valid
  ) %>%
  # add in some random missingness
  r_na(prob=.4)

```

Let's look at the relationship between Height and Age.

```{r}

ggplot(data = df,
       aes(x = Height,
           y = Age)) +
  geom_point()

```

We get a little message, warning us about the missing values.

We can instead use the `geom_point_missing` to display the missing data

```{r}

library(ggmissing)

ggplot(data = df,
       aes(x = Height,
           y = Age)) +
  geom_missing_point()

```

`geom_missing_point()` has (albeit somewhat clumsily) shifted the missing values to now be 10% below the minimum value. However, the missing values need to be shown in a different colour so that missingness becomes preattentive. This code is currently being incorported into `geom_missing_point()`

```{r}
df %>%
  mutate(miss_cat = miss_cat(., "Height", "Age")) %>%
  ggplot(data = .,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age),
           colour = miss_cat)) + 
  geom_point() 
```

However, this is a little verbose.

Examples of the current work, in a stream-of-consciousness style can be seen in the vignette. The functions in this package are basically utility functions, `shadow_shift`, which shifts missing values to 10% below minimum, `shadow_df`  creates a shadow matrix, `miss_cat` creates a new column of missingness status that allows for the plot to create a factor out of missingness. `miss_cat` uses the utility function `shadow_cat`.

# Missing data tidying functions

`ggmissing` uses some missingness transformation functions to set up tables for visualisation.  

The `summarise_missingness` function takes a `data.frame` and then returns a dataframe containing the percentages of missing data, and lists of dataframes containing tally and summary information for the variables and cases. 

```{r}

s_miss <- summarise_missingness(airquality)

s_miss

# overall % missing data
s_miss$percent_missing_df

# % of variables that contain missing data
s_miss$percent_missing_var

# % of cases that contain missing data
s_miss$percent_missing_case

# tabulations of missing data across cases
s_miss$table_missing_case

# tabulations of missing data across variables
s_miss$table_missing_var

# summary information (counts, percentrages) of missing data for variables and cases
s_miss$summary_missing_var
s_miss$summary_missing_case

```

Each of these functions can also be called invididually

```{r}

# overall percentage of missing data
percent_missing_df(airquality)

# % of variables that contain missing data
percent_missing_var(airquality)

# % of cases that contain missing data
percent_missing_case(airquality)

# tabulations of missing data across cases
table_missing_case(airquality)

# tabulations of missing data across variables
table_missing_var(airquality)

# summary information (counts, percentrages) of missing data for variables and cases
summary_missing_var(airquality)
summary_missing_case(airquality)


```

# Other plotting functions

These tables could then be used in these plots

## gg_missing_var

```{r}

gg_missing_var(airquality)

```


## gg_missing_case

```{r}

gg_missing_case(airquality)

```

## gg_missing_which

This shows whether

```{r}

gg_missing_which(airquality)

```

## heatmap of missing data?

I recommend the use of the `vis_miss` function from the [`visdat`](github.com/njtierney/visdat) package.

```{r}

library(visdat)

vis_miss(airquality)

```


# Future Work

`ggmissing` has not seen much attention for the past 6 months or so, and so will be undergoing more changes over the next 6 months.

We will be considering how to display missing data with other visualisations, such as:

- 1D, univaritae distribution plots
- Categorical variables
- Bivariate plots: Scatterplots, Density overlays.
