---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/njtierney/ggmissing?branch=master&svg=true)](https://ci.appveyor.com/project/njtierney/ggmissing)

<!-- [![Travis-CI Build Status](https://travis-ci.org/njtierney/ggmissing.svg?branch=master)](https://travis-ci.org/njtierney/ggmissing) -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# ggmissing

`ggmissing` adds ggplot `geom`s to display missingness. 

## Why?

Missing data is ubiquitous in data analysis. However, vis  points are omitted in ggplot, and although it does provides a warning message telling you that you have missing data, it is tricky to visualise missing data.

`ggmissing` is part of a larger plan for a set of tidy-verse packages focussing on how to tidy, transform, visualise, model, and communicate missing data.

It is still very much under development, and may have unknown bugs, due to the fact that ggplot was not initially built to handle missing data in this way. We will see more active development over the next 6 months.

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

## What does it do?

`ggmissing` provides:

1. Missing data geoms for ggplot (`geom_missing_point`)

2. Tidy summaries of missing data (`summarise_missingness` and friends)

3. Standard plots for exploring missing data (`gg_missing_var`, `gg_missing_case`, `gg_missing_which`)

## Using ggmissing {#ggplot}

### How does it work?

Plotting missing data might sound a little strange - how do you visualise something that is not there? In the past, GGobi and Manet have provided methods of visualising missingness, with one approach being to replace "NA" values with values 10% lower than the minimum value in that variable.

To illustrate, let's explore the relationship between Ozone and Solar radiation from the airquality dataset.

```{r messages = FALSE, error = FALSE}
library(ggmissing)
# devtools::install_github("njtierney/ggmissing")
library(ggplot2)
library(dplyr)

```

```{r}

ggplot(data = airquality,
       aes(x = Ozone,
           y = Solar.R)) +
  geom_point()

```

We get a little message, warning us about the missing values.

We can instead use the `geom_missing_point()` to display the missing data

```{r}

library(ggmissing)

ggplot(data = airquality,
       aes(x = Ozone,
           y = Solar.R)) +
  geom_missing_point()

```

`geom_missing_point()` has shifted the missing values to now be 10% below the minimum value. The missing values are a different colour so that missingness becomes preattentive.

This plays nicely with other parts of ggplot, like faceting - we can split the facet by month:

```{r}
p1 <-
ggplot(data = airquality,
       aes(x = Ozone,
           y = Solar.R)) + 
  geom_missing_point() + 
  facet_wrap(~Month, ncol = 2) + 
  theme(legend.position = "bottom")

p1

```

And then change the theme, just like you do with any other ggplot graphic

```{r}

p1 + theme_bw()  

```




# Missing data tidying functions {#tidying}

`ggmissing` uses some missingness transformation functions to set up tables for visualisation.  

```{r}

# overall percentage of missing data
percent_missing_df(airquality)

# % of variables that contain missing data
percent_missing_var(airquality)

# % of cases that contain missing data
percent_missing_case(airquality)

# tabulations of missing data across cases
table_missing_case(airquality)

# tabulations of missing data across variables
table_missing_var(airquality)

# summary information (counts, percentrages) of missing data for variables and cases
summary_missing_var(airquality)
summary_missing_case(airquality)


```


Each of these functions can also be called all together using `summarise_missingness`, which takes a `data.frame` and then returns a nested dataframe containing the percentages of missing data, and lists of dataframes containing tally and summary information for the variables and cases. 

```{r}

s_miss <- summarise_missingness(airquality)

s_miss

# overall % missing data
s_miss$percent_missing_df

# % of variables that contain missing data
s_miss$percent_missing_var

# % of cases that contain missing data
s_miss$percent_missing_case

# tabulations of missing data across cases
s_miss$table_missing_case

# tabulations of missing data across variables
s_miss$table_missing_var

# summary information (counts, percentrages) of missing data for variables and cases
s_miss$summary_missing_var
s_miss$summary_missing_case

```


# Other plotting functions {#extraplots}

These dataframes from the tidying functions are then used in these plots

## gg_missing_var

```{r}

gg_missing_var(airquality)

```


## gg_missing_case

```{r}

gg_missing_case(airquality)

```

## gg_missing_which

This shows whether

```{r}

gg_missing_which(airquality)

```

## heatmap of missing data?

I recommend the use of the `vis_miss` function from the [`visdat`](github.com/njtierney/visdat) package.

```{r}

# devtools::install_github("njtierney/visdat")
library(visdat)
vis_miss(airquality)

```

# Future Work

`ggmissing` has not seen much attention for the past 6 months or so, and so will be undergoing more changes over the next 6 months, with plans to have the package in CRAN before the end of 2016. 

As such, we plan to extend the `geom_missing` family to include:

- 1D, univariate distribution plots
- Categorical variables
- Bivariate plots: Scatterplots, Density overlays.

There are other utility functions:

- `shadow_shift`, which shifts missing values to 10% below minimum
