---
title: "ggmissing: expanding ggplot to handle missing data"
author: "Nicholas Tierney"
date: "`r Sys.Date()`"
output: 
  html_document: 
    theme: readable
---

```{r setup}

knitr::opts_chunk$set(message = F)

```

This is a draft document/vignette that gets ggplot to display missingness into a plot. 


```{r}
library(dplyr)
library(wakefield)
df <- 
  r_data_frame(
  n = 30,
  id,
  race,
  age,
  sex,
  hour,
  iq,
  height,
  died,
  Scoring = rnorm,
  Smoker = valid
  ) %>%
  r_na(prob=.4)

```


```{r}
library(ggplot2)

ggplot(data = df,
       aes(x = Height,
           y = Age)) + 
  geom_point()

# idea was the plot the missing data as 10% below the minimum value for that variable.

df %>%
  # make missing values 10% below the minimum value for that variable
  mutate(Height = ifelse(is.na(Height), 
                         yes = min(Height, na.rm = T)*0.9,
                         no = Height),
         Age = ifelse(is.na(Age),
                      yes = min(Age, na.rm = T)*0.9,
                      no = Age)) %>%
           ggplot(data = .,
                  aes(x = Height,
                      y = Age)) + 
           geom_point()

```


```{r}

is.na.data.frame(df)

df.shadow <- as.data.frame(is.na.data.frame(df))

# make a function for creating a true/false shadow matrix
shadow_df <- function(x){
  x %>%
    is.na.data.frame %>%
    as.data.frame %>% 
    as_data_frame
}

# remember that TRUE = missing
shadow_df(df)


# Make a window function that transforms missing values to be 10% below the minimum value for that variable
shadow_shift <- function(x){
  ifelse(is.na(x), 
         yes = min(x, na.rm = T)*0.9,
         no = x)  
       # min() might change to something related to the data range
       # possibly use range() to determine the shadow shift
       # Need to also add some jitter/noise to these points to seperate out repeats of the same value
  # for factors, need to add another level (smaller than smallest)
  # need to think about how time is handled as well.
}


df %>%
  # make missing values 10% below the minimum value for that variable
  mutate(Height = shadow_shift(Height),
         Age = shadow_shift(Age)) %>%
  ggplot(data = .,
         aes(x = Height,
             y = Age)) + 
  geom_point()

# OK, so it turns out that I can just shadow shift the data INSIDE ggplot.

ggplot(data = df,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age))) + 
  geom_point()

# now we just need to add in some colour to these points, so that missing data sorta takes upon this "preattentive" phase.

# let's make a new dataset, that is basically contains only the shifted data

df.test <- 
df %>%
  mutate(Height = shadow_shift(Height)) %>% 
  # filter out those observations that are greater than the min
  filter(Height < (min(Height, na.rm = T)*1.1))

df.test.2 <- 
df %>%
  mutate(Age = shadow_shift(Age)) %>% 
  # filter out those observations that are greater than the min
  filter(Age < (min(Age, na.rm = T)*1.1))

# to manage the different quantities of the variables, I could put them into a list, or something that allows me to have a "ragged" dataset

ggplot(data = df,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age))) + 
  geom_point() + 
  geom_point(data = df.test,
             aes(x = Height),
             colour = "Red") + 
  geom_point(data = df.test.2,
             aes(y = Age),
             colour = "Red")


```

Colour by `interaction`, which creates all the different combinations of levels of factors.

```{r}

miss_cat <- 
  df %>% 
  shadow_df %>% 
  select(Height, Age) %>% 
  interaction

miss_cat

```

We can then use this as a colour for missingness

```{r}

# we can then colour by the missingness
df %>%
  mutate(miss_cat = miss_cat) %>% 
  ggplot(data = .,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age),
           colour = miss_cat)) + 
  geom_point() 

```

Getting closer now - we can now just filter the appropriate categories out, as we want TRUE.FALSE, TRUE.TRUE, and FALSE.TRUE to be the same level, we can then colour by missingness.

So we have this variable `miss_cat`, which we add to the data.

We can then use an ifelse statement to re-organise the factor levels.


```{r}

# we can then colour by the missingness
df %>%
  mutate(miss_cat = miss_cat) %>% 
  mutate(miss_cat = ifelse(miss_cat == "TRUE.FALSE" | 
                           miss_cat == "TRUE.TRUE" | 
                           miss_cat == "FALSE.TRUE",
                           yes = "Missing",
                           no = "Not Missing")) %>%
  ggplot(data = .,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age),
           colour = miss_cat)) + 
  geom_point() 

```

Ta da! It works. Not quite as elegant as I'd like, but we're getting there.

Instead of having an ifelse each time we want to do this, let's generalize this to a function that reorganizes the miss_cat based upon the levels of missingness - `shadow_cat`

```{r}

shadow_cat <- function(x){
  ifelse(x == "TRUE.FALSE" | 
         x == "TRUE.TRUE" | 
         x == "FALSE.TRUE",
         yes = "Missing",
         no = "Not Missing")
}

df %>%
  mutate(miss_cat = miss_cat) %>% 
  mutate(miss_cat = shadow_cat(miss_cat)) %>%
  ggplot(data = .,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age),
           colour = miss_cat)) + 
  geom_point() 

```

We shouldn't have to create miss_cat to get this to work, so let's make it into a function, which takes the data, and the variables we want, and performs interaction on them.

```{r}

miss_cat <- function(df, var1, var2){
  df %>% 
    shadow_df %>% 
    select(one_of(var1, var2)) %>% 
    interaction
}

df %>%
  mutate(miss_cat = miss_cat(., "Height", "Age")) %>% 
  mutate(miss_cat = shadow_cat(miss_cat)) %>%
  ggplot(data = .,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age),
           colour = miss_cat)) + 
  geom_point() 

```

But now we have this sorta redundant step of creating it using `miss_cat`, then refactoring it using `shadow_cat`. Really though, we can just put `shadow_cat` inside of `miss_cat`, to reduce coding.

```{r}


miss_cat <- function(df, var1, var2){
  df %>% 
    # make the data into a true/false data frame
    shadow_df %>% 
    # choose the variables of interest
    select(one_of(var1, var2)) %>% 
    # get all the combinations of the levels as factors
    interaction %>%
    # combine them into something sensible for our purposes
    shadow_cat
}

df %>%
  mutate(miss_cat = miss_cat(., "Height", "Age")) %>% 
  ggplot(data = .,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age),
           colour = miss_cat)) + 
  geom_point() 


```

Miles had a couple of ideas, where we could combine all of the information within the data frame into one column, effectively encoding the missingness into a single column. However, one problem with this is that if columns are rearranged or removed, then this single column becomes incorrect. 

In terms of things to do from here, I'd really want to create the same sort of plot as above, using some syntax like:

```
ggplot(data = df,
       aes(x = Height,
           y = Age)) + 
  geom_point(missing = T) 

```

So I guess one possible way forward would be to create a function inside `geom_point()`, called `missing`, that is set to FALSE on default, but when TRUE, it munches the appropriate data and produces the plot.

#Stat approach

Created a new stat `stat_missing_point()` that we use with `geom_missing_point()`.

```{r}

ggplot(data = df,
       aes(x = Height,
           y = Age)) + 
  geom_missing_point()

```

# looking at histograms

```{r}
# displaying missingness for univariate histograms:
# this is quite a different routine compared to what I would be doing normally
# which is to say something like the bind_shadows methods, or something
library(dplyr)
library(ggplot2)
library(naniar)

ggplot(data = airquality,
       aes(x = Ozone)) + 
  geom_histogram()

airquality %>%
  select(Ozone) %>%
  mutate(miss_label = label_missing_1d(Ozone)) %>%
  ggplot(data = .,
         aes(x = shadow_shift(Ozone),
             fill = miss_label)) + 
  geom_histogram(binwidth = 1)

# although, this now brings us to a strange question: how should we consider binning for missing values? And how will missing data be affected differently by different bins?

```

