---
title: "Replacing values with NA"
author: "Nicholas Tierney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replacing values with NA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When you are dealing with missing values, you might want to replace values with
a missing values (NA). This is useful in cases when you know the origin of the
data and can be certain which values should be missing. For example, you might
know that all values of "N/A", "N A", and "Not Available", or -99, or -1 are
supposed to be missing.

`naniar` provides functions to specifically work on this type of problem using
the function `replace_with_na`. This function is the compliment to
`tidyr::replace_na`, which replaces an NA value with a specified value, whereas
`naniar::replace_with_na` replaces a value with an NA:

- `tidyr::replace_na`: Missing values turns into a value (NA --> -99)
- `naniar::replace_with_na`: Value becomes a missing value (-99 --> NA)

In this vignette, we describe some simple use cases for these functions and describe how they work.

# Example data

First, we introduce a small fictional dataset, `df`, which contains some common
features of a dataset with the sorts of missing values we might encounter. This
includes multiple specifications of missing values, such as "N/A", "N A", and
"Not Available". And also some common numeric codes, like -98, -99, and -1.

```{r create-df}

df <- tibble::tribble(
  ~name,           ~x,  ~y,              ~z,  
  "N/A",           1,   "N/A",           -100, 
  "N A",           3,   "NOt available", -99,
  "N / A",         NA,  29,              -98,
  "Not Available", -99, 25,              -101,
  "John Smith",    -98, 28,              -1)

```

## Using `replace_with_na`

What if we want to replace the value -99 in the x column with a missing value?

First, let's load `naniar`:

```{r load-naniar}
library(naniar)
```

Now, we specify the fact that we want to replace -99 with a missing value. To do
so we use the `replace` argument, and specify a named list, which contains the 
names of the variable and the value it would take to replace with NA. 

```{r replace-with-na-ex1}
df %>% replace_with_na(replace = list(x = -99))
```

And say we want to replace -98 as well?

```{r replace-with-na-ex2}
df %>%
  replace_with_na(replace = list(x = c(-99, -98)))
```

And then what if we want to replace -99 and -98 in all the numeric columns, 
x and z?

```{r replace-with-na-ex3}
df %>%
  replace_with_na(replace = list(x = c(-99,-98),
                             z = c(-99, -98)))
```

Using `replace_with_na` works well when we know the exact value to be replaced, 
and for which variables we want to replace, providing there are not many 
variables. But what do you do when you've got many variables you want to observe?

## Extending `replace_with_na`

Sometimes you have many of the same value that you want to replace. For example,
-99 and -98 above, and also the variants of "NA", such as "N/A", and "N / A",
and "Not Available". You might also have certain variables that you want to be
affected by these rules, or you might have more complex rules, like, "only affect variables that are numeric, or character, with this rule".

To account for these cases we have borrowed from [`dplyr`'s scoped variants](http://dplyr.tidyverse.org/reference/scoped.html) and created the 
functions:

- `replace_with_na_all()` Replaces NA for all variables. 
- `replace_with_na_at()` Replaces NA on a subset of variables specified with
character quotes (e.g., c("var1", "var2")). 
- `replace_with_na_if()` Replaces NA based on applying an operation on the 
subset of variables for which a predicate function (is.numeric, is.character, etc) returns TRUE.

Below we will now consider some very simple examples of the use of these functions, so that you can better understand how to use them.

## Using `replace_with_na_all`

Use `replace_with_na_all` when you want to replace ALL values that meet a
condition across an entire dataset. The syntax here is a little different, and
follows the rules for rlang's expression of simple functions. This means that
the function starts with `~`, and when referencing a variable, you use `.x`.

For example, if we want to replace all cases of -99 in our dataset, we write:

```{r replace-with-na-all-ex1}

df %>% replace_with_na_all(condition = ~.x == -99)

```

Likewise, if you have a set of (annoying) repeating strings like various 
spellings of "NA", then I suggest you first lay out all the offending cases:

```{r replace-with-na-all-ex2}

# write out all the offending strings
na_strings <- c("NA", "N A", "N / A", "N/A", "N/ A", "Not Available", "NOt available")
```

Then you write `~.x %in% na_strings` - which reads as "does this value occur
in the list of NA strings".

```{r replace-with-na-all-ex3}

df %>%
  replace_with_na_all(condition = ~.x %in% na_strings)

```

### `replace_with_na_at`

This is similar to `_all`, but instead in this case you can specify the 
variables that you want affected by the rule that you state. This is useful in 
cases where you want to specify a rule that only affects a selected number of 
variables.

```{r replace-with-na-at-ex1}

df %>% 
  replace_with_na_at(.vars = c("x","z"),
                     condition = ~.x == -99)

```

Although you can achieve this with regular `replace_with_na`, it is more concise
to use, `replace_with_na_at`. Additionally, you can specify rules as function,
for example, make a value NA if the exponent of that number is less than 1:

```{r replace-with-na-at-ex2}

df %>% 
  replace_with_na_at(.vars = c("x","z"),
                     condition = ~ exp(.x) < 1)
```

### `replace_with_na_if`

There may be some cases where you can identify variables based on some test 
- `is.character` - are they character variables? `is.numeric` - Are they numeric or double? and a given value inside that type of data. For example, 

```{r replace-with-na-if-ex1}

df %>%
  replace_with_na_if(.predicate = is.character,
                     condition = ~.x %in% ("N/A"))

# or
df %>%
  replace_with_na_if(.predicate = is.character,
                     condition = ~.x %in% (na_strings))

```

This means that you are able to apply a rule to many variables that meet a 
pre-specified condition. This can be of particular use if you have many 
variables and don't want to list them all, and also if you know that there is a
particular problem for variables of a particular class.
