---
title: "Replacing values to NA"
author: "Nicholas Tierney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replacing values to NA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When you are dealing with missing values, you might want to replace values currently not missing to be missing values. This can be useful in cases when you know the origin of the data and can be certain which values should be missing. For example, you might know that all values of "N/A", "N A", and "Not Available", or -99, or -1 are supposed to be missing.

`naniar` provides functions to specifically work on this type of proble using the function `replace_to_na`. This function is the compliment to `tidyr::replace_na`, which replaces a value with an NA value, whereas `naniar::replace_to_na` replaces a value with an NA.

- `tidyr::replace_na`: Missing values turns into a value (NA --> -99)
- `naniar::replace_to_na`: Value becomes a missing value (-99 --> NA)

First, we introduce a sample dataset, `df`

```{r create-df}

df <- tibble::tribble(~name,           ~b,    ~x,  ~y,    ~z,
                      "N/A",           "N/A", 1,   "A",   -100,
                      "N A",           "NOt available",   3,   "N/A", -99,
                      "N / A",         "29",   NA,  NA,    -98,
                      "Not Available", "25",   -99, "E",   -101,
                      "John Smith",    "28",   -98, "F",   -1)

df
```

```{r load-naniar}
library(naniar)
```

So what if we want to replace the value -99 in the x column to be missing?

```{r replace-to-na-ex1}
df %>% replace_to_na(to_na = list(x = -99))
```

OK then, and what if we want to replace -98 as well?

```{r replace-to-na-ex2}
df %>%
  replace_to_na(to_na = list(x = c(-99, -98)))
```

And then what if we want to replace -99 and -98 in all the numeric columns?

```{r replace-to-na-ex3}
df %>%
  replace_to_na(to_na = list(x = c(-99,-98),
                             z = c(-99, -98)))
```

So using `replace_to_na` works well when we know the exact value to be replaced, 
and for which variables we want to replace.

However, sometimes you have many of the same value that you want to replace. 
For example, -99 and -98 above, and also the variants of "NA", such as "N/A", 
and "N / A", and "Not Available". You might also have certain variables that 
you want to be affected by these rules, or you might have more complex rules,
like, "When windspeed is more than 90, Solar Radiation levels are unreliable and
should be marked as missing".

To translate many of these different cases across into a set of functions that 
are easy to remember and use, we have borrowed from [`dplyr`'s scoped variants](http://dplyr.tidyverse.org/reference/scoped.html). This means that:

> - Verbs suffixed with `_all()` apply an operation on all variables.
> - Verbs suffixed with `_at()` apply an operation on a subset of variables specified with character quotes
> - Verbs suffixed with `_if()` apply an operation on the subset of variables for which a predicate function returns TRUE.

This now expands the family to:

- `replace_to_na`
- `replace_to_na_all`
- `replace_to_na_at`
- `replace_to_na_if`

We have also added one more scope, `_where`. 

- `replace_to_na_where`

This allows for complex replacements across variables to be made. This is for cases like "Make a value missing when some other variable meets this condition."

Below we will now consider some very simple examples of the use of these functions, so that you can better understand how to use them.

## Example use of scoped variants of `replace_to_na`

### `replace_to_na_all`

Use `replace_to_all_all` when you want to replace ALL values that meet a condition across an
entire dataset. The syntax here is a little different, and follows the rules for rlang's expression of simple functions. This means that the function starts with `~`, and when referencing a variable, you use `.x`.

For example, if we want to replace all cases of -99 in our dataset, we write:

```{r replace-to-na-all-ex1}

df %>% replace_to_na_all(.funs = ~.x == -99)

```

Likewise, if you have a set of (annoying) repeating strings like various spellings of "NA", then I suggest you do something like the following:

```{r replace-to-na-all-ex1}

na_strings <- c("NA", "N / A", "N/A", "N/ A", "Not Available", "NOt available")

df %>%
  replace_to_na_all(.funs = ~.x %in% na_strings)

```


### `replace_to_na_at`

```{r}

# This only replaces values in Ozone, which are -99. This gives the same result
# as vanilla `replace_to_na` above
airquality %>% replace_to_na_at(.vars = "Ozone",
                                .funs = ~.x == -99)

# However, it can be extended, as you can specify other things that are functions

airquality %>% replace_to_na_at(.vars = "Ozone",
                                .funs = ~ sqrt(.x) < 5)
```

### `replace_to_na_if`

```{r}

# the `if` scoped variant is useful if you know something about all variables
# of a particular type. For example, we might know that all integer columns coded as 1 are actually supposed to be missing.

airquality %>% replace_to_na_if(.predicate = is.integer,
                                .funs = ~.x == 1)

```

### `replace_to_na_where`

```{r}

# unfortunately, this only works when there are no missing values in the
# column already
airquality %>% replace_to_na_where(Ozone = Day > 10)

```

# Notes

- Consider writing a "master" wrapper function that allows the user to call each of these from within th fun - `replace_to_na(method = "if")`, etc.
- might be worthwhile making `replace_to_na` `replace_with_na`.
- ideally with `vars()` you can combine the `select` helpers `contains`, `starts_with`, etc.
- Future features: 
  - Verbs suffixed with `_at()` apply an operation on a subset of variables specified with character quotes quoting function `vars()`. This quoting function accepts select_vars() helpers like starts_with(). Instead of a vars() selection, you can also supply an integerish vector of column positions or a character vector of column names.
  - Verbs suffixed with _if() apply an operation on the subset of variables for which a predicate function returns TRUE. Instead of a predicate function, you can also supply a logical vector.
