---
title: "Replacing values with NA"
author: "Nicholas Tierney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replacing values with NA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When you are dealing with missing values, you might want to replace values
currently not missing with a missing values (NA). This is useful in cases
when you know the origin of the data and can be certain which values should be
missing. For example, you might know that all values of "N/A", "N A", and "Not
Available", or -99, or -1 are supposed to be missing.

`naniar` provides functions to specifically work on this type of proble using
the function `replace_with_na`. This function is the compliment to
`tidyr::replace_na`, which replaces a value with an NA value, whereas
`naniar::replace_with_na` replaces a value with an NA.

- `tidyr::replace_na`: Missing values turns into a value (NA --> -99)
- `naniar::replace_with_na`: Value becomes a missing value (-99 --> NA)

First, we introduce a simple dataset, `df`

```{r create-df}

df <- tibble::tribble(
  ~name,           ~x,  ~y,              ~z,  
  "N/A",           1,   "N/A",           -100, 
  "N A",           3,   "NOt available", -99,
  "N / A",         NA,  29,              -98,
  "Not Available", -99, 25,              -101,
  "John Smith",    -98, 28,              -1)

df <- df %>%
  mutate(date = as.Date(c(-1, 17547, 17549,300000, NA), 
                        origin = "1970-01-01"))
```

```{r load-naniar}
library(naniar)
```

What if we want to replace the value -99 in the x column with a missing value?

```{r replace-with-na-ex1}
df %>% replace_with_na(replace = list(x = -99))
```

OK then, and what if we want to replace -98 as well?

```{r replace-with-na-ex2}
df %>%
  replace_with_na(replace = list(x = c(-99, -98)))
```

And then what if we want to replace -99 and -98 in all the numeric columns, 
x and z?

```{r replace-with-na-ex3}
df %>%
  replace_with_na(replace = list(x = c(-99,-98),
                             z = c(-99, -98)))
```

Using `replace_with_na` works well when we know the exact value to be replaced, 
and for which variables we want to replace.

## Extending replace_with_na

Sometimes you have many of the same value that you want to replace. For example,
-99 and -98 above, and also the variants of "NA", such as "N/A", and "N / A",
and "Not Available". You might also have certain variables that you want to be
affected by these rules, or you might have more complex rules, like, "When
windspeed is more than 90, Solar Radiation levels are unreliable and should be
marked as missing".

To translate many of these different cases across into a set of functions that
are easy to remember and use, we have borrowed from [`dplyr`'s scoped variants](http://dplyr.tidyverse.org/reference/scoped.html). In our case, this
means:

- `replace_with_na_all()` Replaces NA for all variables. 
- `replace_with_na_at()` Replaces NA on a subset of variables specified with
character quotes (e.g., c("var1", "var2")). 
- `replace_with_na_if()` Replaces NA based on applying an operation on the 
subset of variables for which a predicate function returns TRUE.

This now expands the family to:

- `replace_with_na`
- `replace_with_na_all`
- `replace_with_na_at`
- `replace_with_na_if`

We have also added one more scope, `_where`. 

- `replace_with_na_where`

This allows for complex replacements across variables to be made. This is for
cases like "Make a value missing when some other variable meets this condition."

Below we will now consider some very simple examples of the use of these functions, so that you can better understand how to use them.

## Example use of scoped variants of `replace_with_na`

### `replace_with_na_all`

Use `replace_with_na_all` when you want to replace ALL values that meet a
condition across an entire dataset. The syntax here is a little different, and
follows the rules for rlang's expression of simple functions. This means that
the function starts with `~`, and when referencing a variable, you use `.x`.

For example, if we want to replace all cases of -99 in our dataset, we write:

```{r replace-with-na-all-ex1}

df %>% replace_with_na_all(.funs = ~.x == -99)

```

Likewise, if you have a set of (annoying) repeating strings like various 
spellings of "NA", then I suggest you do something like the following:

```{r replace-with-na-all-ex2}

# write out all the offending strings
na_strings <- c("NA", "N / A", "N/A", "N/ A", "Not Available", "NOt available")

df %>%
  replace_with_na_all(.funs = ~.x %in% na_strings)

```


### `replace_with_na_at`

This is similar to `_all`, but instead in this case you can specify the 
variables that you want affected by the rule that you state. This is useful in 
cases where you want to specify a rule that only affects a selected number of 
variables.

```{r replace-with-na-at-ex1}

# This only replaces values in Ozone, which are -99. This gives the same result
# as vanilla `replace_with_na` above
df %>% 
  replace_with_na_at(.vars = c("x","z"),
                     .funs = ~.x == -99)

```

This can be achieved with `replace_with_na`, but it is more concise, and you can
also specify rules as function, for example, make a value NA if the exponent of that number is less than 1

```{r replace-with-na-at-ex2}

df %>% 
  replace_with_na_at(.vars = c("x","z"),
                     .funs = ~ exp(.x) < 1)
```

### `replace_with_na_if`

There may be some cases where you can identify variables based on some test 
- `is.character` - are they character variables? `is.Date` - Are they date type?
and a given value inside that type of data. For example, 

```{r replace-with-na-if-ex1}

df %>%
  replace_with_na_if(.predicate = is.character,
                     .funs = ~.x %in% ("N/A"))

```

### `replace_with_na_where`

```{r}

# unfortunately, this only works when there are no missing values in the
# column already
airquality %>% replace_with_na_where(Ozone = Day > 10)

airquality %>% replace_with_na_where(Ozone = Solar.R > 150)



```

# Notes

- Consider writing a "master" wrapper function that allows the user to call each of these from within th fun - `replace_with_na(method = "if")`, etc.

- Future features: 
  - ideally with `vars()` you can combine the `select` helpers `contains`, `starts_with`, etc.
  - Verbs suffixed with `_at()` apply an operation on a subset of variables specified with character quotes quoting function `vars()`. This quoting function accepts select_vars() helpers like starts_with(). Instead of a vars() selection, you can also supply an integerish vector of column positions or a character vector of column names.
  - Verbs suffixed with _if() apply an operation on the subset of variables for which a predicate function returns TRUE. Instead of a predicate function, you can also supply a logical vector.

Not sure why this doesn't work? 

```r
df %>% 
  replace_with_na_if(.predicate = lubridate::is.Date,
                     .funs = ~.x < as.Date(1, "1970-01-01"))
```

- look at dplyr::recode

```r
: x <- sample(c("a", "b", "c"), 10, replace = TRUE)
x

dplyr::recode(x, a = "Apple")

dplyr::recode(x, a = "Apple", .default = NA_character_)
```
