---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

These are various notes that I want to keep track off online

```{r}
library(simputation)
library(tidyverse)
library(naniar)
library(visdat)

# are there missings in the data?
anyNA(airquality)

# OK, where are these missings?
vis_miss(airquality)
miss_var_summary(airquality)

# OK, let's explore the missings in Solar.R and Ozone

da1 <- impute_lm(airquality, Solar.R ~ .)

as_tibble(airquality)
as_tibble(da1)

vis_miss(da1)

head(da1,3)

```

```{r}

library(simputation)
library(naniar)
library(visdat)
library(tibble)

# in a 

```

You get some data, and it has missing values, which you might identify with a tool like `visdat::vis_miss`

```{r}
library(visdat)
vis_miss(airquality)
```

You might then decide that you are going to impute the values of Solar.R using all available data. The package `simputation` makes this really easy:

```{r}

library(simputation)
da1 <- impute_lm(airquality, Solar.R ~ .)

```

We can look at where the values were imputed:

```{r}

as_tibble(airquality)

as_tibble(da1)

```

But it is not very clear now which values were imputed.

You can do something with `vis_compare`

```{r}
vis_compare(airquality, da1)
```

Unfortunately it picks up on the different class, as Solar.R changed from integer to double. 

To help manage these values there should be a way to store which values were imputed, while remaining firmly in the tidyverse.

I think that something like an `impl` is needed:

> imp_df =  imputation + tbl_df

an `impl` object stores initial missings from when data was imputed, but also updates with the data as the data was changed, but behaves as normal for dplyr functions. It may even have its own dplyr verbs, similar to `sf`  `mutate.impl`, `summarise.impl` etc.

The print method would look really similar to a tibble, with some additional features, it would add braces around the missing values that were imputed, and also state some information about the number of imputed values overall. An idea of what this might look like is shown below

```
#>   # An impl: 153 × 6
#>   # <value> imputated values
#>     Ozone Solar.R  Wind  Temp Month   Day
#>     <int>   <int> <dbl> <int> <int> <int>
#> 1     41     190   7.4    67     5     1
#> 2     36     118   8.0    72     5     2
#> 3     12     149  12.6    74     5     3
#> 4     18     313  11.5    62     5     4
#> 5    {21}   {312}  14.3    56     5     5
#> 6     28    {109}  14.9    66     5     6
#> 7     23     299   8.6    65     5     7
#> 8     19      99  13.8    59     5     8
#> 9      8      19  20.1    61     5     9
#> 10    NA     194   8.6    69     5    10
#> # ... with 143 more rows
#> # ... and  <value> more imputed values
```

Now, this object may then play well with `dplyr` verbs, perhaps an `impl` object will then allow for numerical summaries of missing values to be completed as normal, and might for example behave differently for `summarise` functions in `dplyr`, where something like (sans imputation)

```{r}
airquality %>%
  group_by(Month) %>%
  summarise(mean_ozone = mean(Ozone, na.rm = TRUE))
```

It may provide something different with and `impl` objects, giving summaries about the updated imputated values

```
#> # A tibble: 5 × 2
#>   Month mean_ozone mean_ozone_imp
#>   <int>      <dbl>      <dbl>
#> 1     5   23.61538   <value>
#> 2     6   29.44444   <value>
#> 3     7   59.11538   <value>
#> 4     8   59.96154   <value>
#> 5     9   31.44828   <value>
```

This may also have implications further down the track with ggplot.

Related to this topic, is how these values will be stored, and how imputation will occur in `naniar`.


# stat_impute would then take a imple object, and highlight/differentiate the
# missing values from ggplot

# Imputation in naniar ---------------------------------------------------------

# the `imputation()` or `na_impute` function.
# this would take a dataframe, and then some function for imputing values
# naniar "freezes" the dataframe input, and takes a record of the missing values
# and stores them as a sparse matrix (to save space)
# and then imputes the values.

# so, perhaps something like this:

# na_impute(.data = ,
#           .fun = )

# airquality_imp <- 
  # na_impute(.data = airquality,
  #           .fun = impute_cart(y ~ .))




```{r}

# How do people generally keep track of their imputations in the data?


mutate.nibble <- function(x){
  
  x <- mutate(x)
  # update the shadow index
  shadow_index <- which_na(x)
}

# using the special hidden form of Shadow df

ggplot(data = bind_shadow(airquality),
             aes(x = Temp)) + 
  geom_histogram() + 
  facet_wrap(~Ozone_NA,
             ncol = 1)

ggplot(data = bind_shadow(airquality),
             aes(x = Temp,
                 colour = Ozone_NA)) + 
  geom_density() 

ggplot(data = airquality,
       aes(x = Temp,
           colour = is.na(Ozone))) + 
  geom_density() 

# so you have:
# NA = Missing originally
# !NA = Not missing
# ?NA = Imputed
# However, this means that we can no longer store as TRUE/FALSE, but rather
# as a true factor, with defined kinds of missing values
# 

ggplot(data = airquality_nibble,
       aes(x = Temp,
           colour = Ozone)) + 
  geom_density() 

# imputation + tbl_df = impl_df
# impl object stores initial missings from when data was imputed.
# it would also update with the data as the data was changed
# so would require dplyr verbs with methods mutate.imple, etc.

# stat_impute would then take a imple object, and highlight/differentiate the
# missing values from ggplot

# Imputation in naniar ---------------------------------------------------------

# the `imputation()` or `na_impute` function.
# this would take a dataframe, and then some function for imputing values
# naniar "freezes" the dataframe input, and takes a record of the missing values
# and stores them as a sparse matrix (to save space)
# and then imputes the values.

# so, perhaps something like this:

# na_impute(.data = ,
#           .fun = )

# airquality_imp <- 
  # na_impute(.data = airquality,
  #           .fun = impute_cart(y ~ .))

# and provide a print method VERY similar to tibble
# but it would add braces around the missing values that were imputed
#>   # A {n}ibble: 153 × 6
#>   # <value> imputated values
#>     Ozone Solar.R  Wind  Temp Month   Day
#>     <int>   <int> <dbl> <int> <int> <int>
#> 1     41     190   7.4    67     5     1
#> 2     36     118   8.0    72     5     2
#> 3     12     149  12.6    74     5     3
#> 4     18     313  11.5    62     5     4
#> 5    {21}   {312}  14.3    56     5     5
#> 6     28    {109}  14.9    66     5     6
#> 7     23     299   8.6    65     5     7
#> 8     19      99  13.8    59     5     8
#> 9      8      19  20.1    61     5     9
#> 10    NA     194   8.6    69     5    10
#> # ... with 143 more rows
#> # ... with 

# What do we do when we want to plot these values and colour by imputation?

ggplot(airquality_imp,
       aes(x = Ozone,
           y = Solar.R,
           colour = .imputed)) + 
  geom_impute_point()

# if we would like to do this, we need to have a long form "behind the scenes"

#> 
#> # A {n}ibble: 918 × 3
#> # With <value> imputated values
#> rows      var   value .imputed
#> <int>    <chr> <lgl>    <dbl>
#>   1    1 Ozone   41     FALSE
#> 2      2 Ozone   36     FALSE
#> 3      3 Ozone   12     FALSE
#> 4      4 Ozone   18     FALSE
#> 5      5 Ozone  {21}    TRUE
#> 6      6 Ozone   28     FALSE
#> 7      7 Ozone   23     FALSE
#> 8      8 Ozone   19     FALSE
#> 9      9 Ozone    8     FALSE
#> 10    10 Ozone  NA      FALSE
#> # ... with 908 more rows
#> # ... and 

library(pryr)

index_na <- function(x) which(is.na(x), arr.ind = TRUE)

object_size(index_na(airquality))
```

Another note.

There should be something analogous to `as_shadow` for imputations.

This would results in a dataframe that has the same columns copied (that were imputed), with a suffix _imp.

```{r}
as_shadow(airquality)
```

```{r eval = FALSE}

# these are all just ideas at the moment

is_impute <- function(x) {
  if (length(x) == 0) {
    stop("Input is of length 0, please check your inputs.", call. = FALSE)
    } else{
      # this bit here is where we do the computation of the "frozen" data that existed prior to imputation. This is sorta two steps - was it missing in the original data, and did the imputation process impute the data?
      # maybe I can just try getting this working for doing some really simple mean imputation process.
  factor(is.na(x),
         levels = c(FALSE, TRUE),
         labels = c("!imp", "imp"))
    }
}


add_impute <- function(data){

  if(is.data.frame(data) == FALSE){
    stop("Input must be a data.frame", call. = FALSE)
  }

  data_impute <- purrr::map_df(data, is_impute)

  names(data_shadow) <- paste0(names(data),"_NA")

  data_shadow

}

```

I need some way to track the imputations.

So let's just try using mean imputation.

```{r}

# save a copy of the original positions
dat_shadow <- as_shadow(airquality)

# impute the data.
  
  # what are the values that need to be imputed?
  is.na(airquality)
  
  airquality %>% purrr::map_df(mean,na.rm = TRUE)
  
  # need something that takes a predicate that states "this is numeric AND missing"
  # so that I can then impute this value.
  airquality %>%
    mutate_if(.predicate = case_when(
    ))
  
  airquality %>%
    mutate_if(.predicate = ,
              .funs = function(x) mean(x))
  
  
miss_var_summary(airquality)

is.numeric && is.na ~ mean(.)

```

