---
title: "State of Missing data in CRAN"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
```

Missing data is an annoying and pervasive problem in datascience. Consequently, there are a lot of ways to handle missing data, and a lot of packages on CRAN that help with this.

I have only heard of a couple of main packages for missing data and imputation in R, these were:

- VIM / VIMGUI
- mice
- mi
- Amelia
- imputeTS
- simputation
- MissingDataGUI

So I asked people on twitter what they use: 

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Do <a href="https://twitter.com/hashtag/rstats?src=hash">#rstats</a> people use/prefer one of the following packages for missing data imputation/exploration? Reply for different package! :)</p>&mdash; Nicholas Tierney (@nj_tierney) <a href="https://twitter.com/nj_tierney/status/871607459965673472">June 5, 2017</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

From 43 people:

- 43% used mice
- 7% used mi
- 19% used Amelia
- 11% simputation

I also found a few other examples I had not come across:

- Hmisc::aregImpute
- vtreat

However, this is a small sample. 

Can we explore how many are downloaded?

[Julia Silge](https://juliasilge.com/blog/mining-cran-description/) wrote a really nice [blog post](https://juliasilge.com/blog/mining-cran-description/) that looked at all of the CRAN package description files and did some really nifty text analysis. To do this, she used the function `tools::CRAN_package_db()` to get a list of the R packages available on CRAN.

```{r}
# super handy code from Julia Silge's blog:
# https://juliasilge.com/blog/mining-cran-description/

cran <- tools::CRAN_package_db()

# the returned data frame has two columns with the same name???
cran <- cran[,-65]

# make it a tibble
cran <- tbl_df(cran)

# I want to find which packages mention the words "missing data" in their description.

cran_miss <- cran %>%
  select(Package,
         Description) %>% 
  mutate(has_missing = grepl("imput | miss", Description)) %>%
  filter(has_missing)

```

So now we have a dataframe of the R packages that contain "imput" or "miss" in their description file. 

There's probably a lot of packages in there that we aren't actually interested in, but for the moment, lets see what is popular, by seeing which are downloaded. We can get a count of the number of downloads every month using `cranlongs`.

```{r}

library(cranlogs)

# use cranlogs to find how often packages are downloaded
cran_miss_download <- cran_downloads(packages = cran_miss$Package,
               when = "last-month") %>%
  as_tibble() %>%
  group_by(package) %>%
  mutate(n_dl = sum(count)) %>%
  select(package, n_dl) %>%
  ungroup() %>%
  distinct() %>%
  arrange(-n_dl)

```

We can then do a bar plot of the top 20 packages:

```{r}

cran_miss_download %>%
  slice(1:20) %>%
  ggplot(aes(x = reorder(package, n_dl),
             y = n_dl)) + 
  geom_col() +
  coord_flip()
```

OK, what if we filter out Hmisc, Purrr, gtools, and R.methodsS3

```{r}

cran_miss_download %>%
  slice(5:25) %>%
  ggplot(aes(x = reorder(package, n_dl),
             y = n_dl)) + 
  geom_col() +
  coord_flip()

```

Next, I will filter down the packages used, and find what is good, and what is bad.
