---
title: "Mechanics of shadow_shift in naniar"
author: "Nicholas Tierney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}

knitr::opts_chunk$set(message = F)

```

This vignette describes the mechanics behind `shadow_shift`, and steps you through how the functions work. It is not intended to be a user guide on how to use naniar - for that please read the [README on GitHub](https://github.com/njtierney/naniar). Instead, it serves to document my initial process for creating the functions

We first create some data using the `wakefield` package

```{r}
library(dplyr)
library(wakefield)
set.seed(1531)
df <- 
  r_data_frame(
  n = 30,
  id,
  race,
  age,
  sex,
  hour,
  iq,
  height,
  died,
  Scoring = rnorm,
  Smoker = valid
  ) %>%
  r_na(prob=.4)

```

Exploring the data, we see that ggplot2 omits data points.

```{r}
library(ggplot2)

ggplot(data = df,
       aes(x = Height,
           y = Age)) + 
  geom_point()

```

The missing values are ommitted. This isn't great. One idea is to plot the missing data as 10% below the minimum value for that variable.

```{r}

df %>%
  # make missing values 10% below the minimum value for that variable
  mutate(Height = ifelse(is.na(Height), 
                         yes = min(Height, na.rm = T)*0.9,
                         no = Height),
         Age = ifelse(is.na(Age),
                      yes = min(Age, na.rm = T)*0.9,
                      no = Age)) %>%
           ggplot(data = .,
                  aes(x = Height,
                      y = Age)) + 
           geom_point()

```


So we can make a window function that transforms missing values to be 10% below the minimum value for a variable

```{r}

# is.na.data.frame(df)

df.shadow <- as.data.frame(is.na.data.frame(df))

# make a function for creating a true/false shadow matrix
shadow_df <- function(x){
  x %>%
    is.na.data.frame %>%
    as.data.frame %>% 
    as_data_frame
}

# remember that TRUE = missing
shadow_df(df)

# Make a window function that transforms missing values to be 10% below the minimum value for that variable
shadow_shift <- function(x){
  ifelse(is.na(x), 
         yes = min(x, na.rm = T)*0.9,
         no = x)  
       # min() might change to something related to the data range
       # possibly use range() to determine the shadow shift
       # Need to also add some jitter/noise to these points to seperate out repeats of the same value
  # for factors, need to add another level (smaller than smallest)
  # need to think about how time is handled as well.
}

```

Which means we can do something like this

```{r}

  # make missing values 10% below the minimum value for that variable
ggplot(data = df,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age))) + 
  geom_point()

```


However, we now we just need to add in some colour to these points, so that missing data  takes upon this "pre-attentive" phase. We make a new dataset, that is basically contains only the shifted data.


```{r}

df.test <- 
df %>%
  mutate(Height = shadow_shift(Height)) %>% 
  # filter out those observations that are greater than the min
  filter(Height < (min(Height, na.rm = T)*1.1))

df.test.2 <- 
df %>%
  mutate(Age = shadow_shift(Age)) %>% 
  # filter out those observations that are greater than the min
  filter(Age < (min(Age, na.rm = T)*1.1))

```



```{r}

ggplot(data = df,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age))) + 
  geom_point() + 
  geom_point(data = df.test,
             aes(x = Height),
             colour = "Red") + 
  geom_point(data = df.test.2,
             aes(y = Age),
             colour = "Red")


```

Colour by `interaction`, which creates all the different combinations of levels of factors.

```{r}

miss_cat <- 
  df %>% 
  shadow_df %>% 
  select(Height, Age) %>% 
  interaction

miss_cat

```

We can then use this as a colour for missingness

```{r}

# we can then colour by the missingness
df %>%
  mutate(miss_cat = miss_cat) %>% 
  ggplot(data = .,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age),
           colour = miss_cat)) + 
  geom_point() 

```

Getting closer now - we can now just filter the appropriate categories out, as we want TRUE.FALSE, TRUE.TRUE, and FALSE.TRUE to be the same level, we can then colour by missingness.

So we have this variable `miss_cat`, which we add to the data.

We can then use an ifelse statement to re-organise the factor levels.

```{r}

# we can then colour by the missingness
df %>%
  mutate(miss_cat = miss_cat) %>% 
  mutate(miss_cat = ifelse(miss_cat == "TRUE.FALSE" | 
                           miss_cat == "TRUE.TRUE" | 
                           miss_cat == "FALSE.TRUE",
                           yes = "Missing",
                           no = "Not Missing")) %>%
  ggplot(data = .,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age),
           colour = miss_cat)) + 
  geom_point() 

```

Instead of having an ifelse each time we want to do this, let's generalize this to a function that reorganizes the miss_cat based upon the levels of missingness - `shadow_cat`

```{r}

shadow_cat <- function(x){
  ifelse(x == "TRUE.FALSE" | 
         x == "TRUE.TRUE" | 
         x == "FALSE.TRUE",
         yes = "Missing",
         no = "Not Missing")
}

df %>%
  mutate(miss_cat = miss_cat) %>% 
  mutate(miss_cat = shadow_cat(miss_cat)) %>%
  ggplot(data = .,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age),
           colour = miss_cat)) + 
  geom_point() 

```

We shouldn't have to create miss_cat to get this to work, so let's make it into a function, which takes the data, and the variables we want, and performs interaction on them.

```{r}

miss_cat <- function(df, var1, var2){
  df %>% 
    shadow_df %>% 
    select(one_of(var1, var2)) %>% 
    interaction
}

df %>%
  mutate(miss_cat = miss_cat(., "Height", "Age")) %>% 
  mutate(miss_cat = shadow_cat(miss_cat)) %>%
  ggplot(data = .,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age),
           colour = miss_cat)) + 
  geom_point() 

```

But now we have this  redundant step of creating it using `miss_cat`, then refactoring it using `shadow_cat`. Really though, we can just put `shadow_cat` inside of `miss_cat`, to reduce coding.

```{r}


miss_cat <- function(df, var1, var2){
  df %>% 
    # make the data into a true/false data frame
    shadow_df %>% 
    # choose the variables of interest
    select(one_of(var1, var2)) %>% 
    # get all the combinations of the levels as factors
    interaction %>%
    # combine them into something sensible for our purposes
    shadow_cat
}

df %>%
  mutate(miss_cat = miss_cat(., "Height", "Age")) %>% 
  ggplot(data = .,
       aes(x = shadow_shift(Height),
           y = shadow_shift(Age),
           colour = miss_cat)) + 
  geom_point() 


```

Miles had a couple of ideas, where we could combine all of the information within the data frame into one column, effectively encoding the missingness into a single column. However, one problem with this is that if columns are rearranged or removed, then this single column becomes incorrect.

# The mechanics of shadows


One approach used in programs such as MANET, and ggobi, is to shift missing values below the minimum value, so that they can be plotted on the same axis. We can demonstrate shift these values using a function called `shadow_shift`.


```{r}

library(dplyr)
airquality %>%
  select(Ozone) %>%
  head() %>%
  mutate(air_temp_shift = shadow_shift(Ozone))

```

However, doing this for each variable, so that you can plot this information might involve some code like this:

```{r echo = TRUE}
library(ggplot2)
library(naniar)
ggplot(data = airquality,
       aes(x = shadow_shift(Ozone),
           y = shadow_shift(Solar.R))) +
   geom_point()

```

But then you need to keep track of the missing values, so you can go away and write a function to do this:

```{r}
library(dplyr)
airquality %>%
  mutate(miss_cat = miss_cat(., "Ozone", "Solar.R")) %>%
  ggplot(aes(x = shadow_shift(Ozone),
             y = shadow_shift(Solar.R),
             colour = miss_cat)) +
   geom_point()
```

Which is rather involved, and not ideal, because you are no longer trying to explore missing data, you have now spent time working out how to correctly shift missing values, keep track of what was missing, and then plot this, then add colour, and identify which values are missing.

To make this simpler, you use the ggplot2 geom `geom_miss_point()`. This `geom` allows for missing values to be displayed, and also works with the rest of ggplot2 - themes, and facets as well.

**What is imputed, stays imputed**

**it's invisible!!**


# A look at histograms

```{r}
# displaying missingness for univariate histograms:
# this is quite a different routine compared to what I would be doing normally
# which is to say something like the bind_shadows methods, or something
library(dplyr)
library(ggplot2)
library(naniar)

ggplot(data = airquality,
       aes(x = Ozone)) + 
  geom_histogram()

airquality %>%
  select(Ozone) %>%
  mutate(miss_label = label_missing_1d(Ozone)) %>%
  ggplot(data = .,
         aes(x = shadow_shift(Ozone),
             fill = miss_label)) + 
  geom_histogram(binwidth = 1)

# although, this now brings us to a strange question: how should we consider binning for missing values? And how will missing data be affected differently by different bins?

```

