---
title: "Gallery of missing data visualisations with naniar + visdat"
author: "Nicholas Tierney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gallery of missing data visualisations with naniar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

There are a variety of different plots becoming available in the naniar package. In this vignette, we aim to discuss why you would want to use the visualisations in naniar, and also how to interpret the visualisations created. If you would like to know more about the philosophy of the `naniar` package, you should read the vignette "Getting Started with naniar", parts of this visualisation vignette are be similar/identical to this vignette.

# Getting started

One of the first plots that I recommend you start with when you are first exploring your missing data, is the `vis_miss` plot, which is re-exported from [`visdat`](https://www.github.com/ropensci/visdat).

This plot provides a specified visualiation of the amount of missingd data

# Exploring Missingness Mechanisms

You will then want to eplore different missing data mechanisms. To do this we provide many different plots to assist with this.

The first incorporates the method of shifting missing values so that they can be visualised on the same axes as the regular values, and then colours the missing and not missing points.

# `geom_miss_point()`

```{r}

# using regular geom_point()
ggplot(airquality,
       aes(x = Ozone,
           y = Solar.R)) +
geom_point()

# using  geom_miss_point()
ggplot(airquality,
       aes(x = Ozone,
           y = Solar.R)) +
 geom_miss_point()

# Facets!
ggplot(airquality,
       aes(x = Ozone,
           y = Solar.R)) +
 geom_miss_point() + 
 facet_wrap(~Month)

# Themes
ggplot(airquality,
       aes(x = Ozone,
           y = Solar.R)) +
 geom_miss_point() + 
 theme_dark()


```

# `gg_miss_var()`

This plot shows the number of missing values in each variable.

```{r}

gg_miss_var(airquality)
library(ggplot2)
gg_miss_var(airquality) + labs(y = "Look at all the missing ones")

```


# `gg_miss_case()`

This plot shows the number of missing values in each case.

```{r}
gg_miss_case(airquality)
library(ggplot2)
gg_miss_case(airquality) + labs(x = "Number of Cases")
```


# `gg_miss_fct()`

This plot shows the number of missings in each column, broken down by a categorical variable from the dataset.

```{r}

gg_miss_fct(x = riskfactors, fct = marital)
library(ggplot2)
gg_miss_fct(x = riskfactors, fct = marital) + labs(title = "NA in Risk Factors and Marital status")

```

# `gg_miss_span()`

This plot shows the number of missings in a given span, or breaksize.

```{r}

miss_var_span(pedestrian, hourly_counts, span_every = 3000)
library(ggplot2)
gg_miss_span(pedestrian, hourly_counts, span_every = 3000)
# works with the rest of ggplot
gg_miss_span(pedestrian, hourly_counts, span_every = 3000) + labs(x = "custom")
gg_miss_span(pedestrian, hourly_counts, span_every = 3000) + theme_dark()

```


# `gg_miss_case_cumsum()`

This plot shows the cumulative sum of missing values, reading the rows from the top to bottom.

```{r}
gg_miss_case_cumsum(airquality)
library(ggplot2)
gg_miss_case_cumsum(riskfactors, breaks = 50) + theme_bw()
```

# `gg_miss_var_cumsum()`

This plot shows the cumulative sum of missing values, reading columns from the left to the right of your dataframe.

```{r}
gg_miss_var_cumsum(airquality)
```


# `gg_miss_which()`

This plot shows a set of rectangles that indicate whether there is a missing element in a column or not. 

```{r}
gg_miss_which(airquality)
```

